\thetitle

### 2017-07-03 introduction.tex
convert terminology

\section{Introduction --- Mathematical languages as designed and maintained languages}
Mathematical research is aided by the development of languages that help 
researchers to investigate their subject efficiently. Such a development is an
ongoing process, both for informal and formal languages, that may have both
mathematical and meta-mathematical aspects.

@@

An essentially algebraic theory can be built step by step in the following way.
First, one fixes the set $S$ of sorts. Next, one fixes a set $\Sigma_0$ of total
operations.

\subsection*{Setting up the essentially algebraic theory of E-systems}
The theory of E-systems we present in the present article is an iteration in the
development of theory of type
dependency. The goal has been to describe an essentially algebraic theory of
type dependency that is in close correspondence to the type theory of
\cite{TheBook}, while at the same time it is as economic as possible in the
demand of external resources (such as the natural numbers). Instead of explaning
what a dependent type is via an infinite hierarchy of universes $\UU_0$,
$\UU_1$, \ldots, we set out to describe type families using only the three sorts
that typically occur in type theoretical expressions: contexts, families and 
terms. Although an infinite hierarchy of universes allows one to consider everything
to be a term (because every type is a term of some specified universe), and 
therefore use the literally same mechanisms for handling type expressions as for
term expressions in the process of compilation and verification of a 
construction, this specific setup requires the natural numbers to exist in the 
meta-theory.

@@

An E-system is therefore an algebra consisting of three sorts: $C$ for contexts,
$F$ for families, and $T$ for terms. There is a map 
$\eft:F\to C$ which determines for each family, in which context it is a
family; and there is a map $\ebd:T\to F$ which determines for each term, of
which family it is a term. Thus, the backbones of an E-system can be represented
as
\begin{equation*}
\begin{tikzcd}
T \arrow[d,"\ebd"] \\ F \arrow[d,"\eft"] \\ C
\end{tikzcd}
\end{equation*}
This structure is represented by the following set of rules governing judgmental
equality. First, judgmental equality is postulated to be an equivalence relation
in all three cases (contexts, families and terms), so we require
%\bgroup\small
\begin{align*}
& \inference
  { \jctx{\Gamma}
    }
  { \jctxeq{\Gamma}{\Gamma}
    } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      }
    { \jctxeq{\Delta}{\Gamma}
      } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jctxeq{\Delta}{\greek{E}}
      }
    { \jctxeq{\Gamma}{\greek{E}}
      }
    \\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{A}{A}
    } 
& & \inference
    { \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Gamma}{B}{A}
      }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jfameq{\Gamma}{B}{C}
      }
    { \jfameq{\Gamma}{A}{C}
      }
    \\
& \inference
  { \jterm{\Gamma}{A}{x}
    }
  { \jtermeq{\Gamma}{A}{x}{x}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Gamma}{A}{y}{x}
      }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{y}
      \jtermeq{\Gamma}{A}{y}{z}
      }
    { \jtermeq{\Gamma}{A}{x}{z}
      }
\end{align*}
%\egroup
Second, there are convertibility rules assuring that the notion of being a
family in a certain context, or a term of a family, is compatible with judgmental
equality.
\begin{align*}
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jfam{\Gamma}{A}
    }
  { \jfam{\Delta}{A}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Delta}{A}{B}
      }
    \\
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jterm{\Gamma}{A}{x}
    }
  { \jterm{\Delta}{A}{x}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Delta}{A}{x}{y}
      }
    \\
& \inference
  { \jfameq{\Gamma}{A}{B}
    \jterm{\Gamma}{A}{x}
    }
  { \jterm{\Gamma}{B}{x}
    }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Gamma}{B}{x}{y}
      }
\end{align*}
When $A$ is a family in context $\Gamma$, there is also a context $\ctxext{\Gamma}{A}$,
called the \emph{context extension of $A$}. Context extension allows us to consider
`families over families', where a family $P$ over $A$ in context $\Gamma$, would
simply be a family $P$ over $\ctxext{\Gamma}{A}$. Context extension is represented
by another function $\cftctxext : F\to C$.

@@

Thus, an E-system $\mathbf{E}$ will have an underlying category, which may be called the
category of families of $\mathbf{E}$. The structure of type dependency will by
built on top of this category. This structure will consist of an E-homomorphism
$W_A$ called \emph{weakening by $A$}, for every family $A$ in context $\Gamma$,
an E-homomorphism $S_x$ called \emph{substitution by $x$}, for every term
$x$ of $A$ in context $\Gamma$, and units $\idtm{A}\in T(W_A(A))$ for every
family $A$ in context $\Gamma$, satisfying a handful of further axioms. 
The weakening and substitution homomorphisms
will preserve the extension structure we have discussed above; in other words,
they will be functors preserving the terminal object.

\subsection*{The pre-categorical nature of E-systems}
Although we start our definition of E-systems with a category with term structure,
we do not set out to have a categorical notion of E-systems, in the sense that the structure of
being an E-system can be transported along equivalences of categories. The goal
in this paper is to describe the algebraic structures belonging to an essentially
algebraic theory of type dependency, and this structure turns out to be pre-categorical
rather than categorical. The fact that we've encountered a category at the outset
of our definition of E-systems is in a sense coincidence, rather than born out
of a desire to do type theory categorically and we would consider E-systems to
be pre-categorical models of type theory rather than categorical models of type
theory. The difference between pre-categories and categories is that equality
of objects in a pre-category is considered to be (strict) equality, whereas
equality of objects in a category is considered to be isomorphism of objects,
and we have chosen to use the former notion of equality.

@@

Categories with families are models of a dependent type theory with a
category of contexts, for each context $\Gamma$ a sort $\mathrm{Ty}(\Gamma)$ of
types in context $\Gamma$, and for each type $A\in\mathrm{Ty}(\Gamma)$ a sort
$\mathrm{Tm}(A)$ of terms of type $A$ in context $\Gamma$. We will see that
the sorts $\mathrm{Tm}(A)$ are superfluous, because it will always be 
naturally isomorphic to the set of sections of the projection morphism
$p_A:\ctxext{\Gamma}{A}\to\Gamma$ in the category of contexts. The evident
functor from categories with families to E-systems will forget that the contexts
form a category, and therefore it will not be faithful.
\item Comprehension categories were introduced by Jacobs in \cite{Jacobs1993}, as
a minimal notion of categorical model in type theory, without really making
precise what that means. 
Comprehension categories are models of a dependent type theory with a
category of context, and for each context $\Gamma$ a category of types. 
\end{itemize}

There is a simple test to quickly figure out some basic differences between different
dependent type theories, asking what kind of structure remains if there are no
types at all. In a contextual category, only the terminal object remains.
However, when the presheaf $\mathrm{Ty}$ of a category with families is assumed
to be empty, or when a category $\cat{E}$ fibered over some category is assumed
to be empty, the underlying category of context remains. We shall see that
E-systems are like contextual categories in this respect: only a singleton can
be an E-system with no types. This complicates a thorough comparison between
categories with families and comprehension categories on one side, and E-systems
on the other. We shall construct functors from the categories of categories with
families and comprehension categories directly preserving type dependency, but
these functors will not be faithful.

@@

\subsection*{Overview and conventions}
In \autoref{sec:esys_defn}, we give the categorical definition of E-system in 
several stages. First we define weakening systems, then projection systems and
substitution systems, and finally we bundle them together into the definition
of E-systems. The structure of the definition of weakening systems is 
prototypcial: we first define the notion of pre-weakening systems, which are
categories that come with a pre-weakening structure; then we define the notion
of pre-weakening homomorphism, which is a functor preserving the weakening
structure and we note that it makes sense to ask for the property that the
functors of which the weakening system consists, are also pre-weakening
homomorphisms; and finally, adding this requirement will be give the definition
of weakening systems. We go through this process for each of the definitions
of weakening systems, projection systems, substitution systems and E-systems.

@@

We would also like to remark that although it appears that we introduce possibly
large families of operations (a weakening operation $W_A$ for each morphism $A$,
for instance), E-systems can actually be defined with finitely many
operations. This becomes clearer when one defines an E-system in an arbitrary
category with finite limits.
\end{rmk}

\subsection*{Acknowledgements}
Several people have been indispensible throughout the research process.
I would like to thank 
Richard Garner for the early discussions on the inference rules of type 
dependency we had in Barcelona and in Paris; 
Joachim Kock for the many fruitful discussions and for hosting me at the Universitat
Aut\`onoma de Barcelona during the year 2013-2014;
Steve Awodey for sharing many insights and useful advice which lead to the
definition of an E-system in arbitrary categories with finite limits;
and Vladimir Voevodsky for inviting me to present E-systems at the Homotopy Type
Theory workshop in Oxford (November 2014) and to visit the Institute for Advanced 
Study in Princeton (March 2015), where the definition presented here was
discovered.

>>>>

\section{Introduction}

\begin{comment}
Mathematical research is aided by the development of languages that help 
researchers to investigate their subject efficiently. Such a development is an
ongoing process, both for informal and formal languages, that may have both
mathematical and meta-mathematical aspects.

@@

An essentially algebraic theory can be built step by step in the following way.
First, one fixes the set $S$ of sorts. Next, one fixes a set $\Sigma_0$ of total
operations. 
\end{comment}

\begin{comment}
The theory of E-systems is a theory of type dependency in which we have a notion of family extension. They are called such because if we try to interpret a category with families as a E-systems, one of the things we have to do is to consider lists of dependent types as our notion of families. These are the objects De Bruijn called telescopes. On the other hand, the theory of E-systems is an (essentially) algebraic theory of the notion of structure. A family over a context $\Gamma$ in a E-system is a structure in context $\Gamma$. Therefore I am also considering the name `structure algebras'. Types are a defined notion in E-systems. They are the atomic families (or structures), i.e.~the ones which do not decompose as the extension of two non-trivial structures. 
\end{comment}

\subsection*{Setting up the essentially algebraic theory of E-systems}
The first goal of this paper is to describe an essentially algebraic theory of
type dependency that is in close correspondence to the type theory of
\cite{TheBook}, while at the same time it is as economic as possible in the
demand of requirements of the meta-theory (such as the availability of a type of natural numbers). Instead of explaning
what a dependent type is via an infinite hierarchy of universes $\UU_0$,
$\UU_1$, \ldots, we describe type dependency using only the three sorts
that typically occur in type theoretical expressions: contexts, dependent contexts and 
terms. Although an infinite hierarchy of universes allows one to reduce everything
to a term (because every type is a term of some specified universe), and 
therefore use the literally same mechanisms for handling type expressions as for
term expressions in the process of compilation and verification of a 
construction, this specific setup requires the natural numbers to exist in the 
meta-theory.

@@

a E-system is therefore an algebra consisting of three sorts: $C$ for contexts,
$F$ for families, and $T$ for terms. There is a map 
$\eft:F\to C$ which determines for each family, in which context it is a
family; and there is a map $\ebd:T\to F$ which determines for each term, of
which family it is a term. Thus, the backbones of a E-system can be represented
as
\begin{equation*}
\begin{tikzcd}
T \arrow[d,"\ebd"] \\ F \arrow[d,"\eft"] \\ C
\end{tikzcd}
\end{equation*}
This structure is represented by the following set of rules governing judgmental
equality. First, judgmental equality is postulated to be an equivalence relation
in all three cases (contexts, families and terms), so we require
%\bgroup\small
\begin{align*}
& \inference
  { \jctx{\Gamma}
    }
  { \jctxeq{\Gamma}{\Gamma}
    } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      }
    { \jctxeq{\Delta}{\Gamma}
      } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jctxeq{\Delta}{\greek{E}}
      }
    { \jctxeq{\Gamma}{\greek{E}}
      }
    \\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{A}{A}
    } 
& & \inference
    { \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Gamma}{B}{A}
      }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jfameq{\Gamma}{B}{C}
      }
    { \jfameq{\Gamma}{A}{C}
      }
    \\
& \inference
  { \jterm{\Gamma}{A}{x}
    }
  { \jtermeq{\Gamma}{A}{x}{x}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Gamma}{A}{y}{x}
      }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{y}
      \jtermeq{\Gamma}{A}{y}{z}
      }
    { \jtermeq{\Gamma}{A}{x}{z}
      }
\end{align*}
%\egroup
Second, there are convertibility rules assuring that the notion of being a
family in a certain context, or a term of a family, is compatible with judgmental
equality.
\begin{align*}
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jfam{\Gamma}{A}
    }
  { \jfam{\Delta}{A}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Delta}{A}{B}
      }
    \\
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jterm{\Gamma}{A}{x}
    }
  { \jterm{\Delta}{A}{x}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Delta}{A}{x}{y}
      }
    \\
& \inference
  { \jfameq{\Gamma}{A}{B}
    \jterm{\Gamma}{A}{x}
    }
  { \jterm{\Gamma}{B}{x}
    }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jtermeq{\Gamma}{A}{x}{y}
      }
    { \jtermeq{\Gamma}{B}{x}{y}
      }
\end{align*}
When $A$ is a family in context $\Gamma$, there is also a context $\ctxext{\Gamma}{A}$,
called the \emph{context extension of $A$}. Context extension allows us to consider
`families over families', where a family $P$ over $A$ in context $\Gamma$, would
simply be a family $P$ over $\ctxext{\Gamma}{A}$. Context extension is represented
by another function $\cftctxext : F\to C$.

@@

Thus, a E-system $\mathbf{E}$ will have an underlying category, which may be called the
category of families of $\mathbf{E}$. The structure of type dependency will by
built on top of this category. This structure will consist of a telescopic homomorphism
$W_A$ called \emph{weakening by $A$}, for every family $A$ in context $\Gamma$,
a telescopic homomorphism $S_x$ called \emph{substitution by $x$}, for every term
$x$ of $A$ in context $\Gamma$, and units $\idtm{A}\in T(W_A(A))$ for every
family $A$ in context $\Gamma$, satisfying a handful of further axioms. 
The weakening and substitution homomorphisms
will preserve the extension structure we have discussed above; in other words,
they will be functors preserving the terminal object.

The category of B-systems will be shown to be equivalent to the category of
stratified E-systems. However, we do not arrive at the fact that the functor
$\mathcal{E}:\mathbf{Bsys}\to\mathbf{Ctx}$ is full without imposing an additional
condition on the E-homomorphisms. 
To ensure that the category of stratified E-systems with stratified 
E-homomorphisms between them is a full subcategory of the category 
$\mathbf{Ctx}$, we will require that the underlying functor of each E-homomorphism
lifts factorizations. It is
important to note that this condition is also algebraic, so that the category
of E-systems remains algebraic.

Recall that for any morphism $f:X\to Y$ in a category $\cat{C}$ there is a category
$\mathbf{fact}(f)$ of factorizations of $f$. Also, any functor $F:\cat{C}\to\cat{D}$
determines a functor $F_{\mathbf{fact}(f)}:\mathbf{fact}(f)\to\mathbf{fact}(H(f))$
for any morphism $f$ of $\cat{C}$.

\begin{defn}
A functor $F:\cat{C}\to\cat{D}$ is said to
\define{lift factorizations} if for any $f\in\cat{C}/X$, the functor
$F_{\mathbf{fact}(f)}:\mathbf{fact}(f)\to\mathbf{fact}(F(f))$ is an isomorphism
of categories. 
\end{defn}

\begin{comment}
\begin{rmk}
We may choose the property of lifting factorizations
to involve either an equivalence or an isomorphism of categories. The
isomorphism-version, would say that for every factorization $h'\circ g'$
of $F(f)$ there is a unique factorization $f\jdeq h\circ g$ in $\cat{C}$ such
that $F(g)\jdeq g'$ and $F(h)\jdeq h'$. In the version with equivalences,
the uniqueness is replaced by uniqueness up to isomorphism, and the equalites
are replaced by an isomorphism in $\mathbf{fact}(F(f))$.

It is easier to state the isomorphism version of lifting factorizations with
inference rules and it might be easier to explain this condition on type theoretical
grounds, although the equivalence version has the advantage of being
categorical (i.e.~invariant under equivalence of categories).

In the present
context, it doesn't matter very much which one we pick: the categories in which
we're interested are all posets.
\end{rmk}
\end{comment}

\begin{eg}
For any slice category $\cat{C}/X$, the forgetful functor $\cat{C}/X\to\cat{C}$
lifts factorizations.
\end{eg}

Convention: A functor in this paper is a functor which lifts factorizations.

\subsection*{The pre-categorical nature of E-systems}
Although we start our definition of E-systems with a category with term structure,
we do not set out to have a categorical notion of E-systems, in the sense that the structure of
being a E-system can be transported along equivalences of categories. The goal
in this paper is to describe the algebraic structures belonging to an essentially
algebraic theory of type dependency, and this structure turns out to be pre-categorical
rather than categorical. The fact that we've encountered a category at the outset
of our definition of E-systems is in a sense coincidence, rather than born out
of a desire to do type theory categorically and we would consider E-systems to
be pre-categorical models of type theory rather than categorical models of type
theory. The difference between pre-categories and categories is that equality
of objects in a pre-category is considered to be (strict) equality, whereas
equality of objects in a category is considered to be isomorphism of objects,
and we have chosen to use the former notion of equality.

@@

Categories with families are models of a dependent type theory with a
category of contexts, for each context $\Gamma$ a sort $\mathrm{Ty}(\Gamma)$ of
types in context $\Gamma$, and for each type $A\in\mathrm{Ty}(\Gamma)$ a sort
$\mathrm{Tm}(A)$ of terms of type $A$ in context $\Gamma$. We will see that
the sorts $\mathrm{Tm}(A)$ are superfluous, because it will always be 
naturally isomorphic to the set of sections of the projection morphism
$p_A:\ctxext{\Gamma}{A}\to\Gamma$ in the category of contexts. The evident
functor from categories with families to E-systems will forget that the contexts
form a category, and therefore it will not be faithful.
\item Comprehension categories were introduced by Jacobs in \cite{Jacobs1993}, as
a minimal notion of categorical model in type theory, without really making
precise what that means. 
Comprehension categories are models of a dependent type theory with a
category of context, and for each context $\Gamma$ a category of types.
However, it is not quite clear what kind of theory the theory of comprehension
categories is, which compromises the study of the notion itself.
\end{itemize}

There is a simple test to quickly figure out some basic differences between different
dependent type theories, asking what kind of structure remains if there are no
types at all. In a contextual category, only the terminal object remains.
However, when the presheaf $\mathrm{Ty}$ of a category with families is assumed
to be empty, or when a category $\cat{E}$ fibered over some category is assumed
to be empty, the underlying category of context remains. We shall see that
E-systems are like contextual categories in this respect: only a singleton can
be a E-system with no types. This complicates a thorough comparison between
categories with families and comprehension categories on one side, and E-systems
on the other. We shall construct functors from the categories of categories with
families and comprehension categories directly preserving type dependency, but
these functors will not be faithful.

@@

\subsection*{Overview and conventions}
In \autoref{sec:esys_defn}, we give the categorical definition of E-systems in 
several stages. First we define weakening systems, then projection systems and
substitution systems, and finally we bundle them together into the definition
of E-systems. The structure of the definition of weakening systems is 
prototypcial: we first define the notion of pre-weakening systems, which are
categories that come with a pre-weakening structure; then we define the notion
of pre-weakening homomorphism, which is a functor preserving the weakening
structure and we note that it makes sense to ask for the property that the
functors of which the weakening system consists, are also pre-weakening
homomorphisms; and finally, adding this requirement will be give the definition
of weakening systems. We go through this process for each of the definitions
of weakening systems, projection systems, substitution systems and E-systems.

@@

We would also like to remark that although it appears that we introduce possibly
large families of operations (a weakening operation $W_A$ for each morphism $A$,
for instance), E-systems can actually be defined with finitely many
operations. This becomes clearer when one defines a E-system in an arbitrary
category with finite limits.
\end{rmk}

\subsection*{Acknowledgements}
Several people have been indispensible throughout the research process.
I would like to thank 
Richard Garner for the early discussions on the inference rules of type 
dependency we had in Barcelona and in Paris; 
Joachim Kock for the many fruitful discussions and for hosting me at the Universitat
Aut\`onoma de Barcelona during the year 2013-2014;
Steve Awodey for sharing many insights and useful advice which lead to the
definition of a E-system in arbitrary categories with finite limits;
and Vladimir Voevodsky for inviting me to present E-systems at the Homotopy Type
Theory workshop in Oxford (November 2014) and to visit the Institute for Advanced 
Study in Princeton (March 2015), where the definition presented here was
discovered.

### 2017-07-03 main.tex
convert terminology

\begin{abstract}
We present an algebraic theory of dependent contexts with
three sorts for contexts, dependent contexts and terms, with finitely many operations
between them and finitely many equality axioms. Algebras for this theory are
called context algebras. A distinctive feature of the dependent type theory of
E-systems is that it takes dependent contexts as the notion of `type dependency'

We then compare Voevodsky's B-systems, Dybjer's categories
with families and Jacobs' comprehension categories with context algebras. There is
a full and faithful functor from B-systems to context algebras, of which the image is
the category of stratified context algebras. To have a functor from categories with
families to context algebras, we are using an additonal coherence condition, which
makes sure that the substitution operation of a category with families extends 
to telescopic families. 
\end{abstract}

>>>>

\begin{abstract}
We present an algebraic theory of dependent contexts with
three sorts for contexts, dependent contexts and terms, with finitely many operations
between them and finitely many equality axioms. Algebras for this theory are
called E-systems. A distinctive feature of the dependent type theory of
E-systems is that it takes dependent contexts as the notion of `type dependency'

We then compare Voevodsky's B-systems, Dybjer's categories
with families and Jacobs' comprehension categories with E-systems. There is
a full and faithful functor from B-systems to E-systems, of which the image is
the category of stratified E-systems. To have a functor from categories with
families to E-systems, we are using an additonal coherence condition, which
makes sure that the substitution operation of a category with families extends 
to telescopic families. 
\end{abstract}
